---
title: "R package: queryup"
author: "Guillaume Voisinne"
output: 
  github_document :
    html_preview: true
date: "`r format(Sys.time(), '%Y - %m - %d')`"
---

```{r include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
devtools::load_all()
```

[![R-CMD-check](https://github.com/VoisinneG/queryup/workflows/R-CMD-check/badge.svg)](https://github.com/VoisinneG/queryup/actions)
[![Codecov test coverage](https://codecov.io/gh/VoisinneG/queryup/branch/master/graph/badge.svg)](https://app.codecov.io/gh/VoisinneG/queryup?branch=master)


The `queryup` R package aims to facilitate retrieving information from the UniProt database using R.
Programmatic access to the UniProt database is peformed by submitting queries to the [UniProt website REST API](https://www.uniprot.org/help/api_queries).

## Install

Install the package from github using devtools:

```{r eval=FALSE, message=FALSE, warnings=FALSE}
devtools::install_github("VoisinneG/queryup")
library(queryup)
```

## Queries
Queries combine different fields to identify matching database entries. Here, queries are submitted using the function `query_uniprot()`. In the `queryup` R package, a query must be formattted as a list containing character vectors named after existing UniProt fields. The list of all available fields along with example queries is detailed [here](https://www.uniprot.org/help/query-fields). Different query fields must be matched simultaneously. For instance, the following query uses the fields *gene_exact* to return the UniProt entries of all proteins encoded by gene *Pik3r1* :

```{r message=FALSE, warnings=FALSE}
query <- list("gene_exact" = "Pik3r1")
df <- query_uniprot(query, show_progress = FALSE)
head(df)
```

## Columns

By default, `query_uniprot()` returns a dataframe with  UniProt accession IDs, gene names, organism and Swiss-Prot review status. You can choose which data columns to retrieve using the `columns` parameter. 

```{r message=FALSE, warnings=FALSE}
df <- query_uniprot(query, 
                    columns = c("id", "sequence", "keyword", "gene_primary"),
                    show_progress = FALSE)

```

See this [web page](https://www.uniprot.org/help/return_fields) for all availbale UniProt data columns.
Note that the parameter `columns` and the name of the corresponding column in the output data frame do not necessarily match.

```{r}
names(df)
as.character(df$Sequence[1])
as.character(df$Keywords[1])
```

## Combining query fields

Our first query returned many matches. We can build more specific queries by using more than one query field. By default, matching entries must satisfy all query fields simultaneously. Let's retrieve the only Swiss-Prot reviewed protein entry encoded by gene *Pik3r1* in *Homo sapiens* (taxon: 9606):

```{r message=FALSE, warnings=FALSE}
query <- list( "gene_exact" = "Pik3r1", "reviewed" = "true", "organism_id" = "9606" )
df <- query_uniprot(query, show_progress = FALSE)
print(df)
```

## Multiple items per query field

It is also possible to look for entries that match different items within a single query field. Items from a given query field are looked for independently. Hence, the following query will return all Swiss-Prot reviewed proteins encoded by either *Pik3r1* or *Pik3r2* in either *Mus musculus* (taxon: 10090) or *Homo sapiens* (taxon: 9606): 

```{r message=FALSE, warnings=FALSE}
query <- list( "gene_exact" = c("Pik3r1", "Pik3r2"), "reviewed" = "true", "organism_id" = c("9606", "10090"))
df <- query_uniprot(query, show_progress = FALSE)
print(df)
```

## Queries with invalid entries

If a query containing invalid entries is sent to the UniProt REST API, an error message is returned and no information about the other potentially  valid entries can be retrieved. To overcome this limitation, `queryup` parses the error messages and remove invalid entries from the query. Hence, `query_uniprot()` will return information for valid entries only :

```{r}
invalid_ids <- c("P226", "CON_P22682", "REV_P47941")
valid_ids <- c("A0A0U1ZFN5", "P22682")
ids <- c(invalid_ids, valid_ids)
query <- list("accession_id" = ids)
query_uniprot(query)
```

## Long queries

Because UniProt REST API limits the size of queries, long queries containing more than a few hundreds entries cannot be passed in a single request. To overcome this limitation, the `queryup` package splits long queries into smaller ones. For instance, the dataset `uniprot_entries` that is bundled with the `queryup` package contains information for 1000 UniProt entries. We could retrieve the ENSEMBL ids corresponding to these entries using :

```{r message=FALSE, warning=FALSE}
ids <- uniprot_entries$Entry
query <- list("accession_id" = ids)
columns <- c("gene_names", "xref_ensembl")
df <- query_uniprot(query, columns = columns, show_progress = FALSE)
head(df)
```

## Protein-protein interactions

Another usage could be to retrieve protein-protein interactions amongst a set of UniProt entries:
```{r message=FALSE, warning=FALSE}
ids <- sample(uniprot_entries$Entry, 400)
query <- list("accession_id" = ids, "interactor"= ids)
columns <- "cc_interaction"
df <- query_uniprot(query = query, columns = columns, show_progress = FALSE)
head(df)
```
