---
title: "R package: queryup"
author: "Guillaume Voisinne"
output: 
  github_document :
    html_preview: true
date: "`r format(Sys.time(), '%Y - %m - %d')`"
---

```{r include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

[![Travis-CI Build Status](https://travis-ci.org/VoisinneG/queryup.svg?branch=master)](https://travis-ci.org/VoisinneG/queryup) 


The `queryup` R package aims to facilitate retrieving information from the UniProt database using R.
Programmatic access to the UniProt database is peformed by submitting queries to the [UniProt website REST API](https://www.uniprot.org/help/api_queries).

## Install

Install the package from github using devtools:

```{r message=FALSE, warnings=FALSE}
devtools::install_github("VoisinneG/queryup")
library(queryup)
```

## Queries
Queries combine different fields to identify matching database entries. Here, queries are submitted using the function `query_uniprot()`. In the `queryup` R package, a query must be formattted as a list containing character vectors named after existing UniProt fields. The list of all available fields along with example queries is shown [here](#list-of-available-query fields). Different query fields must be matched simultaneously. For instance, the following query will return the UniProt ids of all proteins encoded by gene *Pik3r1* in *Homo sapiens* (taxon: 9606). 

```{r message=FALSE, warnings=FALSE}
query <- list( "gene_exact" = "Pik3r1", "organism" = "9606" )
df <- query_uniprot(query, columns = c("id", "genes"), show_progress = FALSE)
head(df)
```

## Columns

By default, `query_uniprot()` returns a dataframe with  protein ids, gene names, organism and Swiss-Prot review status. You can choose which data columns to retrieve using the `columns` parameter. 

```{r message=FALSE, warnings=FALSE}
df <- query_uniprot(query, 
                    columns = c("id", "sequence", "keywords"),
                    show_progress = FALSE)

```

See this [section](#list-of-available-data-columns) for all availbale UniProt data columns.
Note that the parameter `column` and the name of the corresponding column in the output dataframe do not necessarily match.

```{r}
names(df)
as.character(df$Sequence[1])
as.character(df$Keywords[1])
```

## Combining query fields

Our first query returned many matches. We can build more specific queries by using more than one query field. By default, matching entries must satisfy all query fields simultaneously. Let's retrieve the only Swiss-Prot reviewed protein entry encoded by gene *Pik3r1* in *Homo sapiens* (taxon: 9606):

```{r message=FALSE, warnings=FALSE}
query <- list( "gene_exact" = "Pik3r1", "reviewed" = "yes", "organism" = "9606" )
df <- query_uniprot(query, show_progress = FALSE)
print(df)
```

## Multiple items per query field

It is also possible to look for entries that match different items within a single query field. Items from a given query field are looked for independently. Hence, the following query will return all Swiss-Prot reviewed proteins encoded by either *Pik3r1* or *Pik3r2* in either *Mus musculus* (taxon: 10090) or *Homo sapiens* (taxon: 9606): 

```{r message=FALSE, warnings=FALSE}
query <- list( "gene_exact" = c("Pik3r1", "Pik3r2"), "reviewed" = "yes", "organism" = c("9606", "10090"))
df <- query_uniprot(query, show_progress = FALSE)
print(df)
```

## List of available query fields

You can view all query fields available using: 

```{r}
list_query_fields()
```

See [here](https://www.uniprot.org/help/query-fields) for a more detailed description and example queries for each field.

## List of available data columns

You can view all data columns retrieveable using : 

```{r}
list_data_columns()
```

Note that the parameter `columns` and the name of the corresponding column in the output dataframe do not necessarily match. See [here](https://www.uniprot.org/help/uniprotkb_column_names) for a more detailed description.
